Changes in accessibility to foodbanks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-22 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
Started the results and discussion and worked on the figures

% Pandoc header
\usepackage[table]{xcolor}

@@

Food insecurity, defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020), has been
associated with reductions in nutritional outcomes (Bhattacharya et al.,
2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999) and physical and
mental health in children and adults (Elgar et al., 2021; Jones, 2017;
Ramsey et al., 2011; Seligman et al., 2010; Stuff et al., 2004). Over at
least the past four decades food banks have become an essential line of
defense against food insecurity in Canadian communities (Black and Seto,
2020; Holmes et al., 2018; Riches, 2002; Tarasuk et al., 2020). In this
respect, Canada is not unlike numerous other wealthy countries where a
systematic dismantling of the welfare state took place in the
intervening period (Tarasuk et al., 2014).

But the emergence of COVID-19, the worst public health crisis since the
1918 flu pandemic, has revealed important social and economic fault
lines, and pre-existing patterns of inequality appear to have been
exacerbated. Along several other dimensions of stress {[}e.g.,
accessibility to health care facilities; Pereira et al. (2021a){]}, this
seems to be the case for food insecurity as well (Laborde et al., 2020).
In the US, for example, it has been estimated that there was an increase
of more than 30\% in household food insecurity, and more than one third
of households were discovered to be newly food insecure - meaning they
did not experience food insecurity before the pandemic (Niles et al.,
2020). In Canada, Men and Tarasuk (2021) report that about 25\% of
individuals who experienced job insecurity (a relatively common
occurrence during the pandemic), also experienced food insecurity.
Similarly, according to Statistics Canada (2020a), in the early stages
of the pandemic almost 15\% of individuals reported living in a
household that faced food insecurity; the risk of food insecurity was
substantially higher for households with children. The difference
between households with and without children was significant, and 11.7\%
of households with children indicated that ``food didn't last and
{[}there was{]} no money to get more'' sometimes or often, compared to
7.3\% of households without children); likewise, 13\% of households with
children indicated that they ``{[}c{]}ouldn't afford balanced meals''
sometimes or often, compared to 8.8\% of households without children.

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While food banks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As recently as 2019, the Hamilton
Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that food banks in Hamilton, Ontario, recorded the highest number
of visitors in the past 29 years; the number of children visiting food
banks (minors up to 18 years old) was 9,125 in March 2019, up from 8,278
the year before.

It is known that the urban food environment, within which people make
their daily food choices, is essential in influencing eating behaviours
and health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility.

The study of place-based geographic accessibility is concerned with
capturing the potential to reach destinations of value using the
transportation network (Páez et al., 2012). Indeed, the Government of
Canada's recent Food Policy has made ``access'' to healthy food a
priority for Canadian communities\footnote{https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042}
and previous survey research has suggested that such accessibility plays
a key role in user satisfaction with food bank service delivery in
Vancouver (Holmes et al., 2018). However, as with research into the
prevalence of food deserts, accessibility to food banks is likely not
evenly distributed throughout a city with variations in access
attributable to transportation network characteristics and the spatial
distribution of food bank locations - as well as the population that
likely needs them. Furthermore, policy responses to the COVID-19
pandemic add to the distress of vulnerable households.
Non-pharmaceutical interventions during the pandemic involving
restrictions in mobility increased the friction of travel, in particular
by transit on which low income populations are more likely to be reliant
(e.g., DeWeese et al., 2020). At the same time, the pandemic has created
additional stress for the operators of foodbanks through disruptions in
the supply chain (e.g., McKay et al., 2021) as well as concerns
surrounding the delivery of service in safe conditions and possible
cancellation of food service programs.

For this study, we aim to look at how the landscape of food banks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, has changed before and during the
pandemic. Have the number of open food bank services diminished? If so,
what was the accessibility to foodbanks before and during the pandemic,
from the perspective of low income households? And finally, who are most
likely to have been impacted by changes to the accessibility landscape?
This paper will first look at the distribution of food banks and related
services before and during the pandemic. Then, we use the balanced
floating catchment area approach of Paez et al. (Paez et al., 2019) to
investigate the accessibility situation. We use a fully disaggregated
approach based on parcel-level data. Socio-economic and demographic data
are drawn from the latest Census of Canada (2016), whereas travel
information is from the most recent regional travel survey from 2016.
This paper follows reproducible research recommendations (see Brunsdon
and Comber, 2020), and the research was conducted using open source
tools for transportation analysis (Lovelace, 2021). The code and data
necessary to reproduce the analysis are available in a public
repository\footnote{add repository}.

\hypertarget{food-insecurity-and-food-bank-use-in-canada}{%
\section{Food Insecurity and Food Bank Use in
Canada}\label{food-insecurity-and-food-bank-use-in-canada}}

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake (Enns
et al., 2020; Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009).
This nutrient deficiency is associated with major health concerns in
Canadians, and particularly those who are at a socioeconomic
disadvantage (Bazerghi et al., 2016). Data on food insecurity in Canada
is collected in several ways. Quantitatively, official government
surveys such as the Household Food Security Survey Module (HFSSM), the
Canadian Community Health Surveys (CCHS), the International Study of
Adults (LISA), and official classifications determined by Health Canada
in relation to sociodemographic variables offer some insight into food
insecurity (Gundersen et al., 2018; Kirkpatrick and Tarasuk, 2008;
Tarasuk and Vogt, 2009). Nationally, it has been found that food
insecurity impacts approximately 12.3\% of Canadian households (Tarasuk
et al., 2014)

Food banks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
foodbanks can vary by region and by country, and these organizations can
include not only prepared meals and aliments, but also shared spaces to
connect in community gardens and community kitchens (Wakefield et al.,
2013). Although in their origin foodbanks were meant to be provide a
temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
evolved into a community practice to secure emergency food supplies for
those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013).

In Canada, the number of food banks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of food banks
and their use comes from the non-profit association Food Banks Canada,
which conducts an annual assessment through its affiliated members. Food
Banks Canada's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member food banks across the
country found that Canadians visited food banks 1.1 million times in
March of 2018. Of those accessing food banks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to Food Banks
Canada's 2018 data, single-adult households represent 45\% of those
utilizing food banks despite making up 28\% of Canada's population, 19\%
are single-parent households (compared to 10\% nationally), and 35\% of
those accessing food bank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing food banks list social or
disability assistance as their primary source of income. Moreover,
children aged 0-18 are over-represented as food bank users, making up
35\% of those accessing food bank services despite representing 20\% of
the country's population. Likewise, using data from the 2011-2012 CCHS,
Tarasuk et al. (2019) found higher odds of food insecurity amongst
households relying on social assistance, those without a university
degree or with children under the age of 18, and individuals that lived
alone, renters, and those identifying as Aboriginal. While surveys
revealed that only 20 to 30 percent of those experiencing food
insecurity were found to frequent food banks in Canada (Tarasuk et al.,
2014), research from Ottawa (Enns et al., 2020) and Vancouver (Black and
Seto, 2020) suggests that long-term users tend to be older, have health
or mobility challenges, live in large households, and less likely to
have employment income.

Currently with the COVID-19 pandemic, disrupted economies, rising
unemployment rates, and changes in poverty levels have disrupted the
food environment and led to higher rates of food insecurity (Niles et
al., 2020). Pandemic-related food security studies in the US have found
a substantial increase in households experiencing food insecurity for
the first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from Food Banks Canada showed that 52\% of member food banks reported an
increase in usage in March of 2020 when initial lockdown restrictions
were put in place across much of Canada. The pandemic also created
significant staffiing issues with 42\% of food banks reporting a
reduction in volunteers. However, 53\% of food banks reported a decrease
in use into the summer of 2020 which food banks attributed to emergency
financial support programs from the federal government were.
Nevertheless, some of these benefit programs were temporary, suggesting
many households may again turn to food bank services to meet their
needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing food banks, the locational or
transportation accessibility aspect of food bank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socioeconomic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For food banks specifically, previous qualitative
research in Ontario has noted difficulties associated with
transportation for trips to the food bank (Smith-Carrier et al., 2017).
For example, Smith et al. (2017) note that participants access food
banks using a variety of modes of transportation with respondent stating
``I wish it {[}the food bank{]} was a little more centrally located.
Because if I didn't have a bike I'd have to walk it all the way out
there and back. I wonder about people who don't.'' From this, the
authors conclude that ``transportation can be challenging, particularly
if the food bank is situated in a remote location.'' To offer greater
insight into the role of transportation and location in food bank
accessibility, this research examines how geographic access to food
banks and food services changed in Hamilton before and during the
COVID-19 pandemic.

@@

For the research in this paper we adopt the balanced floating catchment
area approach of Paez et al. (2019). This method for estimating
accessibility is a form of the common two-stage floating catchment area
method (Luo and Wang, 2003; Radke and Mu, 2000). Floating catchment
areas are used to estimate accessibility when there are potential
congestion effects, and operate by calculating first the \emph{demand}
for spatially distributed services. The demand (usually the number of
people who require a service) is used to calculate a level of service.
In a second step, the level of service is allocated back to the
population. Demand and level of service are allocated using some form of
distance-decay to embody the geographical principle that, given a
choice, people prefer to travel less than more when reaching
destinations.

@@

Floating catchment area methods are prone to overestimation of the
population and the level of service due to multiple-counting. The
population at \(P_i\) is allocated to \emph{every} service point \(j\)
for which \(w_{ij}>0\). Similarly, the level of service at \(LOS_j\) is
allocated to \emph{every} population point for which \(w_{ji}>0\). This
inflation effect has been known for several years, and several
modifications have been proposed to mitigate it (Delamater, 2013; e.g.,
Wan et al., 2012). A definitive solution to this issue was presented by
Paez et al. (2019). In order to avoid the multiple-counting in the
summations, the population and the level of service need to be allocated
\emph{proportionally}. This is achieved by standardizing the weights as
follows: \[
w_{ij}^\text{st} = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
\] \noindent and: \[
w_{ji}^\text{st} = \frac{w_{ji}}{\sum_{j=1}^Jw_{ji}}
\]

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-results-sensitivity-analysis-1} \caption{\label{fig:sensitivity-analysis}Accessibility as a function of threshold}\label{fig:plot-results-sensitivity-analysis}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-levels-of-service-1} \caption{\label{fig:levels-of-service}Levels of service at each facility pre-COVID-19 (top panel) and during COVID-19 (middle panel). The bottom panel shows the change in level of service.}\label{fig:plot-levels-of-service}
\end{figure}

@@

\caption{\label{fig:accessibility}Accessibility by traffic analysis zone pre-COVID-19 (top panel) and during COVID-19 (middle panel). The bottom panel shows the change in accessibility.}\label{fig:plot-accessibility}
\end{figure}

>>>>

% Pandoc header
\usepackage[table]{xcolor}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This is a
condition that has been associated with reductions in nutritional
outcomes (Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008;
Olson, 1999) and physical and mental health in children and adults
(Elgar et al., 2021; Jones, 2017; Ramsey et al., 2011; Seligman et al.,
2010; Stuff et al., 2004). Over at least the past four decades foodbanks
have become an essential line of defense against food insecurity in
Canadian communities (Black and Seto, 2020; Holmes et al., 2018; Riches,
2002; Tarasuk et al., 2020). In this respect, Canada is not unlike
numerous other wealthy countries where a systematic dismantling of the
welfare state took place in the intervening period (Tarasuk et al.,
2014).

But the emergence of COVID-19, the worst public health crisis since the
1918 flu pandemic, has revealed important social and economic fault
lines, and pre-existing patterns of inequality appear to have been
exacerbated. Along several other dimensions of stress (such as
accessibility to health care facilities, see Pereira et al., 2021a),
this seems to be the case for food insecurity as well (Laborde et al.,
2020). In the US, for example, it has been estimated that there was an
increase of more than 30\% in household food insecurity, and more than
one third of households were discovered to be newly food insecure -
meaning they did not experience food insecurity before the pandemic
(Niles et al., 2020). In Canada, Men and Tarasuk (2021) report that
about 25\% of individuals who experienced job insecurity (a relatively
common occurrence during the pandemic), also experienced food
insecurity. Similarly, according to Statistics Canada (2020a), in the
early stages of the pandemic almost 15\% of individuals reported living
in a household that faced food insecurity; the risk of food insecurity
was substantially higher for households with children. The difference
between households with and without children was significant, and 11.7\%
of households with children indicated that ``food didn't last and
{[}there was{]} no money to get more'' sometimes or often, compared to
7.3\% of households without children; likewise, 13\% of households with
children indicated that they ``{[}c{]}ouldn't afford balanced meals''
sometimes or often, compared to 8.8\% of households without children.

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While foodbanks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As recently as 2019, the Hamilton
Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that foodbanks in Hamilton, Ontario, recorded the highest number
of visitors in the past 29 years; the number of children visiting
foodbanks (minors up to 18 years old) was 9,125 in March 2019, up from
8,278 the year before.

It is known that the urban food environment, within which people make
their daily food choices, is essential in influencing eating behaviours
and health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to foodbanks. Although
previous work has explored differences in \emph{accessing} foodbanks,
such as how some households utilize foodbanks over short periods of time
while others regularly utilize foodbanks as longer-term resource (e.g.,
Enns et al., 2020), we are not aware of any research that has focused on
the geographic component of accessibility.

The study of place-based geographic accessibility is concerned with
capturing the potential to reach destinations of value using the
transportation network (Páez et al., 2012). Indeed, the Government of
Canada's recent Food Policy has made ``access'' to healthy food a
priority for Canadian communities\footnote{https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042}
and previous survey research has suggested that such accessibility plays
a key role in user satisfaction with foodbank service delivery in
Vancouver (Holmes et al., 2018). However, as with research into the
prevalence of food deserts, accessibility to foodbanks is unlikely to be
evenly distributed throughout a city due to variations in the
characteristics of the transportation network characteristics, and the
spatial distribution of foodbank locations and the population they
serve. Furthermore, policy responses to the COVID-19 pandemic may well
add to the distress of vulnerable households. Non-pharmaceutical
interventions during the pandemic involving restrictions in mobility
have increased the friction of travel, in particular by transit on which
low income populations are more likely to be reliant (e.g., DeWeese et
al., 2020). At the same time, the pandemic has created additional stress
for the operators of foodbanks through disruptions in the supply chain
(e.g., McKay et al., 2021) as well as concerns surrounding the delivery
of service in safe conditions and possible cancellation of food service
programs.

For this study, we aim to look at how the landscape of foodbanks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Have the
number of open foodbank services diminished? If so, what were the levels
of accessibility to foodbanks before the pandemic from the perspective
of low income households, and how did they change during the pandemic?
And finally, who are most likely to have been impacted by changes to the
accessibility landscape? This paper will first look at the distribution
of foodbanks and related services before and during the pandemic. Then,
we use the balanced floating catchment area approach of Paez et al.
(2019) to investigate the accessibility situation. We use a fully
disaggregated approach based on parcel-level data. Socio-economic and
demographic data are drawn from the latest Census of Canada (2016),
whereas travel information is from the most recent regional travel
survey from 2016. This paper follows reproducible research
recommendations (see Brunsdon and Comber, 2020), and the research was
conducted using open source tools for transportation analysis (Lovelace,
2021). The code and data necessary to reproduce the analysis are
available in a public repository\footnote{add repository}.

\hypertarget{food-insecurity-and-foodbank-use-in-canada}{%
\section{Food Insecurity and foodbank Use in
Canada}\label{food-insecurity-and-foodbank-use-in-canada}}

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to insufficient nutrient intake
(Enns et al., 2020; Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt,
2009). Nutrient deficiency, in turn, is associated with major population
health concerns, and particularly among those who are at a
socio-economic disadvantage (Bazerghi et al., 2016). Data on food
insecurity in Canada is collected in several ways. Quantitatively,
official government surveys such as the Household Food Security Survey
Module (HFSSM), the Canadian Community Health Surveys (CCHS), the
International Study of Adults (LISA), and official classifications
determined by Health Canada in relation to socio-demographic variables
offer some insight into food insecurity (Gundersen et al., 2018;
Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009). Nationally, it
has been found that food insecurity impacts approximately 12.3\% of
Canadian households (Tarasuk et al., 2014)

Foodbanks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
foodbanks can vary by region and by country, and these organizations can
include not only prepared meals and aliments, but also shared spaces to
connect in community gardens and community kitchens (Wakefield et al.,
2013). Although in their origin foodbanks were meant to be provide a
temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
evolved into a community practice to secure emergency food supplies for
those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013).

In Canada, the number of foodbanks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of foodbanks
and their use comes from the non-profit association foodbanks Canada,
which conducts an annual assessment through its affiliated members.
foodbanks Canada's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member foodbanks across the
country found that Canadians visited foodbanks 1.1 million times in
March of 2018. Of those accessing foodbanks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to foodbanks
Canada's 2018 data, single-adult households represent 45\% of those
utilizing foodbanks despite making up 28\% of Canada's population, 19\%
are single-parent households (compared to 10\% nationally), and 35\% of
those accessing foodbank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing foodbanks list social or
disability assistance as their primary source of income. Moreover,
children aged 0-18 are over-represented as foodbank users, making up
35\% of those accessing foodbank services despite representing 20\% of
the country's population. Likewise, using data from the 2011-2012 CCHS,
Tarasuk et al. (2019) found higher odds of food insecurity amongst
households relying on social assistance, those without a university
degree or with children under the age of 18, and individuals that lived
alone, renters, and those identifying as Aboriginal. While surveys
revealed that only 20 to 30 percent of those experiencing food
insecurity were found to frequent foodbanks in Canada (Tarasuk et al.,
2014), research from Ottawa (Enns et al., 2020) and Vancouver (Black and
Seto, 2020) suggests that long-term users tend to be older, have health
or mobility challenges, live in large households, and are less likely to
have employment income.

Currently with the COVID-19 pandemic, disrupted economies, rising
unemployment rates, and changes in poverty levels have disrupted the
food environment and led to higher rates of food insecurity (Niles et
al., 2020). Pandemic-related food security studies in the US have found
a substantial increase in households experiencing food insecurity for
the first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from Foodbanks Canada showed that 52\% of member foodbanks reported an
increase in usage in March of 2020 when initial lockdown restrictions
were put in place across much of Canada. The pandemic also created
significant staffing issues with 42\% of foodbanks reporting a reduction
in volunteers. However, 53\% of foodbanks reported a decrease in use
into the summer of 2020 which foodbanks attributed to emergency
financial support programs from the federal government were.
Nevertheless, some of these benefit programs were temporary, suggesting
many households may again turn to foodbank services to meet their needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing foodbanks, the locational or
transportation accessibility aspect of foodbank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socio-economic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For foodbanks specifically, previous qualitative
research in Ontario has noted difficulties associated with
transportation for trips to the foodbank (Smith-Carrier et al., 2017).
For example, Smith et al. (2017) found that participants access
foodbanks using a variety of modes of transportation with a respondent
stating ``I wish it {[}the foodbank{]} was a little more centrally
located. Because if I didn't have a bike I'd have to walk it all the way
out there and back. I wonder about people who don't.'' From this, the
authors conclude that ``transportation can be challenging, particularly
if the foodbank is situated in a remote location.'' To offer greater
insight into the role of transportation and location in foodbank
accessibility, this research examines how geographic access to foodbanks
and food services changed in Hamilton before and during the COVID-19
pandemic.

@@

For the research in this paper we adopt the balanced floating catchment
area approach of Paez et al. (2019). This method for estimating
accessibility is a form of the common two-stage floating catchment area
method (Luo and Wang, 2003; Radke and Mu, 2000). Floating catchment
areas are used to estimate accessibility when there are potential
congestion effects, and operate by calculating first the \emph{demand}
for spatially distributed services. The demand (e.g., the number of
people who require a service) is used to calculate a level of service.
In a second step, the level of service is allocated back to the
population. Demand and level of service are allocated using some form of
distance-decay to embody the geographical principle that, given a
choice, people prefer to travel less than more when reaching
destinations.

@@

Floating catchment area methods are prone to overestimation of the
population and the level of service due to multiple-counting. The
population at \(P_i\) is allocated to \emph{every} service point \(j\)
for which \(w_{ij}>0\). Similarly, the level of service at \(L_j\) is
allocated to \emph{every} population point for which \(w_{ji}>0\). This
inflation effect has been known for several years, and several
modifications have been proposed to mitigate it (Delamater, 2013; e.g.,
Wan et al., 2012). A definitive solution to this issue was presented by
Paez et al. (2019). In order to avoid the multiple-counting in the
summations, the population and the level of service need to be allocated
\emph{proportionally}. This is achieved by standardizing the weights as
follows: \[
w_{ij}^\text{st} = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
\] \noindent and: \[
w_{ji}^\text{st} = \frac{w_{ji}}{\sum_{j=1}^Jw_{ji}}
\]

@@

Figure \ref{fig:foodbanks} shows the location of the foodbanks in the
City of Hamilton and their status. Before the pandemic there were 58 of
which 14 closed during the pandemic. As seen in the figure, foodbanks
tend to be predominantly located in the central parts of the city. This
is not surprising: population density is high there, and it is also the
part of the city where lower income households are more numerous in
absolute and relative terms (see Figure
\ref{fig:low-income-households}). Alas, this is also where most of the
closures during the pandemic happened.

\begin{figure}
\includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-location-foodbanks-1} \caption{\label{fig:foodbanks}Location of foodbanks and operation status; the dotted box is an inset of the central part of the City of Hamilton}\label{fig:plot-location-foodbanks}
\end{figure}
\begin{figure}
\includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-low-income-households-1} \caption{\label{fig:low-income-households}Number and proportion of households with incomes less than CAD 40,000.}\label{fig:plot-low-income-households}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-results-sensitivity-analysis-1} \caption{\label{fig:sensitivity-analysis}Accessibility as a function of threshold}\label{fig:plot-results-sensitivity-analysis}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-levels-of-service-1} \caption{\label{fig:levels-of-service}Levels of service at each facility pre-COVID-19 (top panel) and during COVID-19 (bottom panel).}\label{fig:plot-levels-of-service}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-levels-of-service-changes-1} \caption{\label{fig:levels-of-service-changes}Changes in levels of service at each facility from pre-COVID-19 to during COVID-19.}\label{fig:plot-levels-of-service-changes}
\end{figure}

@@

\caption{\label{fig:accessibility}Accessibility by traffic analysis zone pre-COVID-19 (top panel) and during COVID-19 (bottom panel).}\label{fig:plot-accessibility}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{Accessibility-Foodbanks-Hamilton_files/figure-latex/plot-accessibility-changes-1}

\caption{\label{fig:accessibility-changes}Changes in accessibility from pre-COVID-19 to during COVID-19.}\label{fig:plot-accessibility-changes}
\end{figure}
